A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-17 tex/BiSig.tex
[ tex ] Fix mistakes in Anno
[ tex ] Minor

\author{Liang-Ting Chen}
\email{liang.ting.chen.tw@gmail.com}
\orcid{0000-0002-3250-1331}
\author{Hsiang-Shang Ko}
\orcid{0000-0002-2439-1048}
\email{joshko@iis.sinica.edu.tw}

\affiliation{%
  \institution{Academia Sinica}
  \streetaddress{128 Academia Road}
  \city{Taipei}
  \country{Taiwan}
  \postcode{115}
}

\title{A Theory of Bidirectional Type Synthesis for Simple Types}

\begin{abstract}
  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
  While progress has been made in developing type-checking techniques on a case-by-case basis, the lack of a general theory poses a difficulty in implementing a robust `type-checker generator' for diverse languages.

  In response to this difficulty, we develop a \emph{simple} yet \emph{general} and \emph{constructive} theory of bidirectional type synthesis where \emph{bidirectional type synthesis} refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
  First, our theory is simple but general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
  Specifically, `the proof' of logical decidability of type synthesis is instantiated to a type synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
  We introduce the notion of bidirectionalisation to refine annotatability in bidirectional typing.
  Type synthesis and bidirectionalisation result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception to locate missing annotations.
  By formalizing our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
\end{abstract}

>>>>

\author{Liang-Ting Chen}
\email{liang.ting.chen.tw@gmail.com}
\orcid{0000-0002-3250-1331}
\author{Hsiang-Shang Ko}
\orcid{0000-0002-2439-1048}
\email{joshko@iis.sinica.edu.tw}
\affiliation{%
  \institution{Academia Sinica}
  \streetaddress{128 Academia Road}
  \city{Taipei}
  \country{Taiwan}
  \postcode{115}
}

\title{A Theory of Bidirectional Type Synthesis for Simple Types}
\begin{abstract}
  Type checking and inference serve as the transition from parsed abstract syntax trees to well-typed programs.
  While progress has been made in developing type-checking techniques on a case-by-case basis, the lack of a general theory poses a difficulty in implementing a robust `type-checker generator' for diverse languages.

  In response to this difficulty, we develop a \emph{simple} yet \emph{general} and \emph{constructive} theory of bidirectional type synthesis where \emph{bidirectional type synthesis} refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
  First, our theory is simple yet general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
  Specifically, the `proof' of logical decidability of type synthesis is instantiated to a synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
  We introduce the notion of bidirectionalisation to refine annotatability in bidirectional typing.
  Type synthesis and bidirectionalisation result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception to locate missing annotations.
  By formalizing our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
\end{abstract}
